The Algorand Virtual Machine

### 2021-03-05 tex/avm.tex
Change how we handle low total stake
Signed-off-by: aybehrouz <behrouz_ayati@yahoo.com>

\usepackage{hyperref}

@@

    To mitigate this problem, for calculating a user's stake at the time step \(t\), instead of using the raw ALGO
    balance, we use the minimum of a \emph{trust value} the system has calculated for the user and the user's
    ALGO balance:
    \[
        Stake_{u,t} = \min (Balance_{u,t}, Trust_{u,t})
    \]
    For estimating the value of \(Trust_{u,t}\) we use the following formula:
    \[
        Trust_{u,t} = \max (M_{u,t}, \beta Balance_{u,t})
    \]
    Where \(M_{u,t}\) is the exponential moving average of the ALGO balance of the user \(u\) at the time step \(t\),
    and \(\beta\) is a constant between \(0\) and \(1\) determining the minimum trust value of a user.

    The agreement protocol, at the time step \(t\), will use \(\sum_{u}Stake_{u,t}\) to determine the required number of
    votes for the confirmation of a block. At the start of the agreement protocol, \(\beta\) is initialized to \(0\).
    If after confirming a block we have:
    \[
        \sum_{u}Stake_u < \gamma \sum_{u} Balance_u
    \]
    The protocol will increase the value of \(\beta\) to make sure that the total stake of the system is not too low.

    If after confirming a block we have \(\beta > 0\) and:
    \[
        \sum_{u}Stake_u > \lambda \sum_{u} Balance_u
    \]
    The protocol will decrease the value of \(\beta\) until it eventually reaches \(0\) again.

    In our system a user who held ALGOs and participated in the consensus for a long time is more trusted
    than a user with a higher balance whose balance has increased recently. An attacker who has obtained a large
    amount of ALGOs, also needs to hold them for a long period of time before being able to attack our system.

    For calculating the exponential moving average of a time series at the time step \(t\), we can use the following
    recursive formula:
    \[
        M_t = (1 - \alpha) M_{t-1} + \alpha X_t = M_{t-1} + \alpha (X_t - M_{t-1})
    \]
    Where:
    \begin{itemize}
        \item The coefficient \(\alpha\) is a constant smoothing factor between \(0\) and\(1\) which represents the
        degree of weighting decrease, A higher \(\alpha\) discounts older observations faster.
        \item \(X_t\) is the value of the time series at the time step \(t\).
        \item \(M_t\) is the value of the EMA at the time step \(t\).
    \end{itemize}

    Usually an account balance will not change in every time step, and we can use older values of EMA for calculating
    \(M_t\):
    \[
        M_t = (1 - \alpha)^{t-k}M_k + [1 - (1 - \alpha)^{t - k}]X
    \]
    Where:
    \[
        X = X_{k+1} = X_{k+2} = \dots = X_t
    \]
    When \(|nx| \ll 1\) we can use the binomial approximation  \((1 + x)^n \approx 1 + nx\) to further simplify this
    formula:
    \[
        M_t = M_k + (t - k) \alpha (X - M_k)
    \]

    For choosing the value of \(\alpha\) we can consider the number of time steps that the trust value of a user needs
    for reaching a specified fraction of his account balance. We know that for large \(n\) and \(|x| < 1\) we have \(
    (1 + x)^n \approx e^{nx}\), so by letting \(M_k = 0\) and
    \(n = t - k\)
    we can write:
    \[
        \alpha =- \frac{\ln\left(1 - \frac{M_{n+k}}{X}\right)}{n}
    \]
    The value of \(\alpha\) for a desired configuration can be calculated by this equation. For instance, we could
    calculate the \(\alpha\) for a relatively good configuration in which \(M_{n+k} = 0.8X\) and \(n\) equals to the
    number of time steps of 10 years.

>>>>

\usepackage{hyperref}
%\binoppenalty=99999
%\relpenalty=99999

@@

    To mitigate this problem, for calculating a user's stake at the time step \(t\), instead of using the raw ALGO
    balance, we use the minimum of a \emph{trust value} the system has calculated for the user and the user's
    ALGO balance:
    \[
        S_{u,t} = \min (B_{u,t}, Trust_{u,t})
    \]
    The agreement protocol, at the time step \(t\), will use \(\sum_{u}S_{u,t}\) to determine the required
    number of votes for the confirmation of a block, and we let \(Trust_{u,t} = M_{u,t}\), where \(M_{u,t}\) is the
    exponential moving average of the ALGO balance of the user \(u\) at the time step \(t\).

    In our system a user who held ALGOs and participated in the consensus for a long time is more trusted
    than a user with a higher balance whose balance has increased recently. An attacker who has obtained a large
    amount of ALGOs, also needs to hold them for a long period of time before being able to attack the system.

    For calculating the exponential moving average of a user's balance at the time step \(t\), we can use the following
    recursive formula:
    \[
        M_{u,t} = (1 - \alpha) M_{u,t-1} + \alpha B_{u,t} = M_{u,t-1} + \alpha (B_{u,t} - M_{u,t-1})
    \]
    Where:
    \begin{itemize}
        \item The coefficient \(\alpha\) is a constant smoothing factor between \(0\) and \(1\) which represents the
        degree of weighting decrease, A higher \(\alpha\) discounts older observations faster.
        \item \(B_{u,t}\) is the balance of the user \(u\) at the time step \(t\).
        \item \(M_{u,t}\) is the EMA for the user \(u\) at the time step \(t\).
    \end{itemize}

    Usually an account balance will not change in every time step, and we can use older values of EMA for calculating
    \(M_{u,t}\): (In the following equations the \(u\) subscript is dropped for simplicity)
    \[
        M_{t} = (1 - \alpha)^{t-k}M_{k} + [1 - (1 - \alpha)^{t - k}]B
    \]
    Where:
    \[
        B = B_{k+1} = B_{k+2} = \dots = B_{t}
    \]
    We know that when \(|nx| \ll 1\) we can use the binomial approximation \({(1 + x)^n \approx 1 + nx}\). So, we can
    further simplify this formula:
    \[
        M_{t} = M_{k} + (t - k) \alpha (B - M_{k})
    \]

    For choosing the value of \(\alpha\) we can consider the number of time steps that the trust value of a user needs
    for reaching a specified fraction of his account balance. We know that for large \(n\) and \(|x| < 1\) we have
    \((1 + x)^n \approx e^{nx}\), so by letting \(M_{u,k} = 0\) and \(n = t - k\) we can write:
    \[
        \alpha =- \frac{\ln\left(1 - \frac{M_{n+k}}{B}\right)}{n}
    \]
    The value of \(\alpha\) for a desired configuration can be calculated by this equation. For instance, we could
    calculate the \(\alpha\) for a relatively good configuration in which \(M_{n+k} = 0.8B\) and \(n\) equals to the
    number of time steps of 10 years.

    In our system a newly created account will not have voting power for some time, no matter how high its
    balance is. While this is a desirable property, in case a large proportion of total system tokens are
    transferred to newly created accounts, it can result in too much voting power for older accounts. This may decrease
    the degree of decentralization in our system.

    However, this situation is easily detectable by comparing the total stake of the system with the total balance of
    users. If after confirming a block the total stake of the system goes too low and we have:
    \[
        \sum_{u}S_{u,t} < \gamma \sum_{u}B_{u,t}
    \]
    The protocol will perform a \emph{time shift} in the system: the time step of the system
    will be incremented for \(m\) steps while no blocks will be confirmed. This will increase the value of \(M_{u,t}\)
    for new accounts with a non-zero balance, giving them more influence in the agreement protocol.

    For calculating the value of \(m\) which determines the amount of time shift in the system, we should note that when
    \(B_{u,t} = B_{u, t-1} = B_u\), we can derive a simple recursive rule for the stake of a user:
    \[
        S_{u,t} = (1 - \alpha) S_{u,t-1} + \alpha B_u
    \]
    Therefore, we have:
    \[
        \sum_{u}S_{u,t} = (1 - \alpha) \sum_{u}S_{u,t - 1} + \alpha \sum_{u}B_u
    \]
    This equation shows that when the balance of users is not changing over time the total stake of the system is the
    exponential average of the total ALGOs of the system. Consequently, when we shift the time for \(m\) steps, we can
    calculate the new total stake of the system from the following equation:

    \[
        \sum_{u}S_{u,t+m} = (1 - \alpha)^{m}\sum_{u}S_{u,t} + [1 - (1 - \alpha)^{m}]\sum_{u}B_u
    \]
    Hence, if we want to increase the total stake of the system from \(\gamma \sum_{u}B_u\) to \(\lambda \sum_{u}B_u\),
    we can obtain \(m\) from the following formula, when \(\alpha\) is small:
    \[
        m = \frac{1}{\alpha} \ln \left(\frac{\gamma - 1}{\lambda - 1}\right)
    \]
